---
title: "Kenya HealthCare"
author: "Esther"
date: "2023-10-01"
output:
  html_document: default
  '''''': default
editor_options: 
  markdown: 
    wrap: 72
---

## Problem Statement

1.  What are the demograpic distribution of surveyed individuals?
2.  What percentage of the surveyed population had health insurance?
3.  Distribution of time since last hospital visit?
4.  Do the patient do routine health checks?(How often)
5.  Have the Individuals taken any cancer screening?
6.  How do various factors e.g. health insurance, hospital visits,cancer
    screening change over time?
7.  Investigate the relationship between monthly household income and
    access to healthcare, insurance coverage, and health outcomes.
8.  Correlations between variables to identify potential associations or
    dependencies.
9.  Find whether there are disparities in healthcare access or outcomes
    based on gender, age, income, or other demographic factors.

```{r echo=FALSE}

library(readxl)
library(dplyr)
library(purrr)
library(ggplot2)
library(tidyr)
library(stringr)
library(forcats)
library(corrplot)
setwd("~/data_analytics/health_care")
Healthcare <- read_excel("~/data_analytics/health_care/Healthcare Dataset.xlsx")
head(Healthcare)
```

```{r}
str(Healthcare)
```

```{r}
summary(Healthcare)
```

```{r}
dim(Healthcare)
```

```{r}
colnames(Healthcare)
```

```{r eval=FALSE, include=FALSE}
Healthcare$Gender <- as.factor(Healthcare$Gender)
Healthcare$`Marital Status` <- as.factor(Healthcare$`Marital Status`)
Healthcare$Age <- as.factor(Healthcare$Age)
Healthcare$`Employment Status` <- as.factor(Healthcare$`Employment Status`)
Healthcare$`Monthly Household Income` <- as.factor(Healthcare$`Monthly Household Income`)
Healthcare$last_check_up_years <- as.numeric(Healthcare$last_check_up_years)
Healthcare$Cancerscreening <- as.logical(Healthcare$Cancerscreening)
Healthcare$Interval_of_screening <- as.numeric(Healthcare$Interval_of_screening)
Healthcare$Insured_last_visit <- as.factor(Healthcare$Insured_last_visit)
Healthcare$routine_checkup <- as.factor(Healthcare$routine_checkup)

```

```{r}
tail(Healthcare)
```

### Rename columns

```{r}
Healthcare <- Healthcare %>%
  rename(Children = 'How many children do you have, if any?', Insured = 'Have you ever had health insurance?', InsuranceName = 'If yes, which insurance cover?', Last_hospital_visit_months = 'When was the last time you visited a hospital for medical treatment? (In Months)' , Insured_last_visit ='Did you have health insurance during your last hospital visit?', routine_checkup = 'Have you ever had a routine check-up with a doctor or healthcare provider?', last_check_up_years = 'If you answered yes to the previous question, what time period (in years) do you stay before having your routine check-up?',  Cancerscreening ='Have you ever had a cancer screening (e.g. mammogram, colonoscopy, etc.)?', Interval_of_screening = 'If you answered yes to the previous question, what time period (in years) do you stay before having your Cancer screening?')
summary(Healthcare)
```

### Identify missing values

```{r}
 Healthcare %>% map(~sum(is.na(.)))
 sum(duplicated(Healthcare))
```

```{r}
Healthcare <- Healthcare %>%
  select (-c(`_tags`, `__version__`, `_submitted_by`, `_status`, `Date and Time`, `_notes`, `_validation_status`,`_submission_time`, `Your Picture_URL`,`_id`, `Your Picture`,`_uuid`, '_index'))
summary(Healthcare)
```

```{r echo=FALSE}
Healthcare$Gender <- as.factor(Healthcare$Gender)
Healthcare$`Marital Status` <- as.factor(Healthcare$`Marital Status`)
Healthcare$Age <- as.factor(Healthcare$Age)
Healthcare$`Employment Status` <- as.factor(Healthcare$`Employment Status`)
Healthcare$`Monthly Household Income` <- as.factor(Healthcare$`Monthly Household Income`)
Healthcare$last_check_up_years <- as.numeric(Healthcare$last_check_up_years)
Healthcare$Cancerscreening <- as.logical(Healthcare$Cancerscreening)
Healthcare$Interval_of_screening <- as.numeric(Healthcare$Interval_of_screening)
Healthcare$Insured_last_visit <- as.factor(Healthcare$Insured_last_visit)
Healthcare$routine_checkup <- as.factor(Healthcare$routine_checkup)
summary(Healthcare)
```

```{r echo=FALSE}
# Calculate the mean of the "Children" variable, excluding NA values
mean_children <- median(Healthcare$Children, na.rm = TRUE)

# Replace missing (NA) values in the "Children" variable with the calculated mean
Healthcare <- Healthcare %>%
  mutate(Children = ifelse(is.na(Children), mean_children, Children))
str(Healthcare)

```

```{r echo=FALSE}
Healthcare_G <- Healthcare %>%
  filter(!is.na(Gender))
Healthcare_G <- Healthcare_G %>%
  group_by(Gender) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Create the bar plot with percentage labels
ggplot(Healthcare_G, aes(Gender, y = percentage,fill = Gender, label = sprintf("%.1f%%", percentage))) +
  geom_col() +
  geom_text(vjust = -0.5) +  # Add data labels above the bars
  labs(
    title = "Gender count",
    x = "",
    y = ""
  ) +
  scale_fill_manual(values = c("Male" = "#27cf9d", "Female" = "purple")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Format y-axis as percentages
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )+
   guides(fill = FALSE)
```

```{r echo=FALSE}
Healthcare$Children[Healthcare$Children >= 14] <- NA
ggplot(Healthcare, aes(Children))+
  geom_histogram(binwidth = 0.5, fill = '#27cf9d')+

labs(
    title = "Number of Children",
    x = "",
    y = ""
  ) +

  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r echo=FALSE}
ggplot(Healthcare, aes(x = Age)) +
  geom_bar( fill = '#27cf9d')+
  labs(
    title = "Age Category",
    x = "",
    y = ""
  ) +

  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
  
 

```

```{r echo=FALSE}
Healthcare_ <- Healthcare %>%
  filter(!is.na(Gender), !is.na(Age))
ggplot(Healthcare_, aes(x = Age, fill = Gender)) +
  geom_bar() +
  labs(
    title = "Population Pyramid",
    x = " ",
    y = "",
    fill = "Gender"
  ) +
  scale_fill_manual(values = c("Male" = "#27cf9d", "Female" = "purple")) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r}
summary(Healthcare)
```

### Marital Status

```{r echo=FALSE}
Healthcare_m <- Healthcare %>%
  group_by(`Marital Status`) %>%
  summarise(Percentage = round((n() / nrow(Healthcare)) * 100, 2)) %>%
  arrange(desc(Percentage))

# Print the resulting summary table
print(Healthcare_m)


```

### Insured

```{r echo=FALSE}

Healthcare_insured <- Healthcare %>%
  group_by(Insured) %>%
  filter(!is.na(Insured)) %>%
  summarise(Percentage = round((n() / nrow(Healthcare)) * 100, 2)) %>%
  arrange(desc(Percentage))
ggplot(Healthcare_insured, aes(x = "", y = "", fill = `Insured`)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(Percentage, "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
                      ggtitle("Percentage Insured")
```

### Age Categories Insured

```{r echo=FALSE}
ggplot(Healthcare_, aes(x = Age, fill = Insured)) +
  geom_bar() +
  labs(title = "Insured vs. Uninsured by Age", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )

```

### Male to Female Insured

```{r}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$Gender) & !is.na(Healthcare$Insured), ]
ggplot(Healthcare_cleaned, aes(x = Gender, fill = Insured)) +
  geom_bar() +
  labs(title = "Insured vs. Uninsured by Gender", y = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

### Type of insurance

```{r}
Healthcare_cover <- Healthcare %>%
  group_by(InsuranceName)%>%
  
  summarise(Count = n()) 
print(Healthcare_cover)
```

### Clean Insurance

```{r echo=FALSE}
expanded_dataset <- Healthcare %>%
  separate_rows(InsuranceName, sep = ",") %>%
  separate_rows(InsuranceName, sep = "\r\n")
expanded_dataset <- expanded_dataset %>%
  mutate(InsuranceName = trimws(str_to_title(InsuranceName)))
insurance_counts <- expanded_dataset %>%
 
  group_by(InsuranceName) %>%
  summarise(Count = n())
insurance_counts


```

```{r echo=FALSE}
expanded_dataset <- expanded_dataset %>%
  mutate(
    InsuranceName = na_if(InsuranceName, "NA"),
    InsuranceName = replace_na(InsuranceName, "Unknown"),
    InsuranceName = case_when(
      grepl("NHIF", InsuranceName, ignore.case = TRUE) ~ "NHIF",
      grepl("NGIF", InsuranceName, ignore.case = TRUE) ~ "NHIF",
      grepl("NHIT", InsuranceName, ignore.case = TRUE) ~ "NHIF",
      grepl("MHIF", InsuranceName, ignore.case = TRUE) ~ "NHIF",
      grepl("HIF", InsuranceName, ignore.case = TRUE) ~ "NHIF",
      grepl("Jubilee", InsuranceName, ignore.case = TRUE) ~ "Jubilee",
      grepl("Jubilee Insurance", InsuranceName, ignore.case = TRUE) ~ "Jubilee",
      grepl("CIC", InsuranceName, ignore.case = TRUE) ~ "CIC",
      grepl("vIC", InsuranceName, ignore.case = TRUE) ~ "CIC",
      grepl("CIC Group Insurance", InsuranceName, ignore.case = TRUE) ~ "CIC",
      grepl("CIC Medical Cover", InsuranceName, ignore.case = TRUE) ~ "CIC",
      grepl("aar", InsuranceName, ignore.case = TRUE) ~ "AAR",
      grepl("ara", InsuranceName, ignore.case = TRUE) ~ "AAR",
      grepl("Absa", InsuranceName, ignore.case = TRUE) ~ "Absa Insurance",
      grepl("Apa", InsuranceName, ignore.case = TRUE) ~ "APA Insurance",
      grepl("uap", InsuranceName, ignore.case = TRUE) ~ "UAP OLD MUTUAL",
         grepl("old Mutual", InsuranceName, ignore.case = TRUE) ~ "UAP OLD MUTUAL",
      grepl("san", InsuranceName, ignore.case = TRUE) ~ "Sanlam",
      grepl("Sam", InsuranceName, ignore.case = TRUE) ~ "Sanlam",
      grepl("Brit", InsuranceName, ignore.case = TRUE) ~ "Britam",
      grepl("Cooperative", InsuranceName, ignore.case = TRUE) ~ "Cooperative Cover",
      grepl("direct", InsuranceName, ignore.case = TRUE) ~ "DirectLine",
      grepl("aar", InsuranceName, ignore.case = TRUE) ~ "AAR",
      grepl("Afya", InsuranceName, ignore.case = TRUE) ~ "Afya Health",
      grepl("kcb", InsuranceName, ignore.case = TRUE) ~ "KCB Insurance",
      grepl("Kenn", InsuranceName, ignore.case = TRUE) ~ "Kenindia",
      grepl("health", InsuranceName, ignore.case = TRUE) ~ "Health Cover",
      grepl("Marte", InsuranceName, ignore.case = TRUE) ~ "Linda Mama",
      grepl("Mate", InsuranceName, ignore.case = TRUE) ~ "Linda Mama",
      grepl("N", InsuranceName, ignore.case = TRUE) ~ "NA",
      grepl("N/A", InsuranceName, ignore.case = TRUE) ~ "NA",
      grepl("No", InsuranceName, ignore.case = TRUE) ~ "NA",
      grepl("Nil", InsuranceName, ignore.case = TRUE) ~ "NA",
      grepl("Null", InsuranceName, ignore.case = TRUE) ~ "NA",
      grepl("NA", InsuranceName, ignore.case = TRUE) ~ "NA",
     !grepl("[a-zA-Z]", InsuranceName) ~ "Unknown",
      TRUE ~ InsuranceName
    )
  )

insurance_counts <- expanded_dataset %>%
  group_by(InsuranceName) %>%
  summarise(Count = n())%>%
  arrange(desc(Count))
insurance_counts

```

```{r echo=FALSE}
threshold <- 10
expanded_dataset <- expanded_dataset %>%
  left_join(insurance_counts, by = "InsuranceName") %>%
  mutate(
    InsuranceName = case_when(
      Count < threshold ~ "Others",
      TRUE ~ as.character(InsuranceName)  # Leave other values unchanged
    )
  ) %>%
  select(-Count) 
insurance_counts <- expanded_dataset %>%
  group_by(InsuranceName) %>%
  summarise(Count = n() ) %>%  
  arrange(desc(Count))

insurance_counts <- insurance_counts %>%
  mutate(Percentage = (Count / sum(Count)) * 100)
insurance_counts

```

### Percentage

```{r echo=FALSE}
insurance_top <- head(insurance_counts)

ggplot(insurance_top, aes(x = fct_reorder(InsuranceName, Percentage), y = Percentage, fill = InsuranceName)) +
  geom_text(aes(label = scales::percent(Percentage / 100)), hjust = -0.1) +
  geom_col() + 
  labs(
    title = "Insurance Distribution", 
    x = "", 
    y = "") + 
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 0),
    legend.position = "none"
  
  ) +
  scale_fill_discrete(guide = guide_legend(title = NULL)) +
  coord_flip()

```

```{r echo=FALSE}
insurance_tail<- tail(insurance_counts,10)

ggplot(insurance_tail, aes(x = fct_reorder(InsuranceName, Percentage), y = Percentage)) +
  geom_col(fill ="#27cf9d") + 
  geom_text(aes(label = scales::percent(Percentage / 100)), vjust = -0.5) + 
  labs(
    title = "Insurance Distribution", 
    x = "", 
    y = "") + 
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(hjust = 0),
    legend.position = "none"
  
  ) +
  scale_fill_discrete(guide = guide_legend(title = NULL)) +
  coord_flip()
```

```{r}
summary(expanded_dataset)
```

```{r echo=FALSE}
ggplot(Healthcare,aes(routine_checkup))+
  geom_bar(fill = "#27cf9d")+
   labs(title = "Routine Checkup", y = "", x ="") +
  theme_minimal()+

  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

### Routine CheckUp vs Insured

```{r echo=FALSE}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$routine_checkup) & !is.na(Healthcare$Insured), ]

ggplot(Healthcare_cleaned, aes(x = routine_checkup, fill = Insured)) +
  geom_bar() +
  labs(title = "Insured vs. Uninsured by Routine Check Up", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
 theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r include=FALSE}
Healthcare_summary <- Healthcare_cleaned %>%
  group_by(routine_checkup, Insured) %>%
  summarise(Count = n(), Percentage = n() / nrow(Healthcare_cleaned) * 100)
Healthcare_summary
```

### Employment

```{r echo=FALSE}
ggplot(Healthcare, aes(`Employment Status` ))+
  geom_bar(fill = "#27cf9d")+
  labs(x = "",
       y = "",
       title = "Employment Status")+
  theme(
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

### Employment Status and Insurance Coverage

```{r echo=FALSE}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$`Employment Status`) & !is.na(Healthcare$Insured), ]

ggplot(Healthcare_cleaned, aes(x = `Employment Status`, fill = Insured)) +
  geom_bar() +
  labs(title = "Insured vs. Uninsured by Employment Status", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
 theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r echo=FALSE}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$`Employment Status`) & !is.na(Healthcare$routine_checkup), ]

ggplot(Healthcare_cleaned, aes(x = `Employment Status`, fill = routine_checkup)) +
  geom_bar() +
  labs(title = "Employement Status vs Routine Check Up", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
 theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r echo=FALSE}
ggplot(Healthcare, aes(`Monthly Household Income` ))+
  geom_bar(fill = "#27cf9d")+
  labs(x = "",
       y = "",
       title = "Monthly Household Income")+
  theme(
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)
     
  ) + 
  coord_flip()
```

```{r echo=FALSE}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$`Monthly Household Income`) & !is.na(Healthcare$routine_checkup), ]

ggplot(Healthcare_cleaned, aes(x = `Monthly Household Income`, fill = routine_checkup)) +
  geom_bar() +
  labs(title = "Household Income vs Routine Check Up", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
 theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{r echo=FALSE}
Healthcare_cleaned <- Healthcare[!is.na(Healthcare$`Monthly Household Income`) & !is.na(Healthcare$Insured), ]


ggplot(Healthcare_cleaned, aes(x = `Monthly Household Income`, fill = Insured)) +
  geom_bar() +
  labs(title = "Household Income Vs Insured", y = "", x = "") +
  theme_minimal()+
  scale_fill_manual(values = c("Yes" = "green", "No" = "purple"))+
 theme(
    legend.position = "right",
    panel.grid.major = element_blank(),  # Hide major grid lines
    panel.grid.minor = element_blank(),  # Hide minor grid lines
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5)  # Center the title
     
  )
```

```{}
```

```{undefined echo=FALSE}

```
